---
title: 'Homework: Numerical linear algebra'
output:
  html_document:
    df_print: paged
  html_notebook:
    code_folding: none
    highlight: textmate
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: {
    Macros: {
      bA: "{\\boldsymbol{A}}",
      bx: "{\\boldsymbol{x}}",
      bb: "{\\boldsymbol{b}}"
    }
  }
});
</script>


```{r setup, include=FALSE}
source(file.path("..", "R", "util.R"))

required_packages <- c("microbenchmark")
install_and_load_packages(required_packages) 
```


# Exercise 1: Comparing different numerical linear algebra algorithms for solving linear systems

In this exercise, we consider the problem of solving a linear system $\bA \bx = \bb$ for $\bx$.
We compare the three methods we learned in the class: LU, Cholesky, and QR decompositions.
(Of course, LU applies to more general systems and QR can be used to solve least squares, but here we focus on positive definite systems.)

## Part A: Racing for solution &mdash; speed comparison 

We first compare their computational speed. 
Fill in the code below and `bench::mark()` the three algorithms.

**Questions:**
What are relative speeds among the algorithms?
Do relative speeds agree with what you expect from the complexity analyses?
If not (quite), why might that be?  
**Answers:** LU costs around 133ms, cholesky decomposition costs around 180ms, QR decomposition costs around 500ms. The relative speeds between cholesky and QR agrees with what I expect from the complexity analysis because QR has a cost of $O(2np^2)$ while cholesky has $O(\frac{1}{3}p^3)$. Hence, when n (sample size) is large, we expect QR to be slower than cholesky. However, the relative speed for LU is not what I expect from the complexity analyses because LU is expected to have a cost of $O(\frac{2}{3}p^3)$. This could be due to the fact that LU decomposition in solve() function uses the LAPACK package which uses a more efficient version of triangular backsolve or forwardsolve than the function I wrote. 

**Note:**
I misspoke about the R's built-in `chol()` function during the lecture:
when applied to a positive-definite `A`, the function actually returns an _upper-triangular_ matrix `R` such that `t(R) %*% R == A`.

```{r, eval=TRUE}
# Import the `rand_positive_def_matrix` function
source(file.path("R", "random_matrix.R"))

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
b <- rnorm(mat_size)

solve_via_cholesky <- function(A, b) {
  L <- t(chol(A))
  x <- backsolve(t(L), forwardsolve(L, b))
  return(x)
}

bench::mark(
  solve(A, b)
)
bench::mark(
  solve_via_cholesky(A, b)
)
bench::mark(
  qr.solve(A, b)
)
```

## Part B: Competition in terms of numerical accuracy/stability

We now compare the three methods in terms of numerical accuracy/stability.
To this end, we set up the following simulation study. 
We first generate a "ground truth" solution vector $\bx_0$.
We then compute an "numerical-error-free" $\bb = \bA \bx_0$ by carrying out the matrix-vector multiplication using the `long double` type, which (on most hardware and compilers) provides [additional 12 bits of precision](https://en.wikipedia.org/wiki/Extended_precision#x86_extended_precision_format).
Of course, the vector $\bb$ computed as such still suffers from numerical errors, but the idea is that the numerical errors from this high-precision matrix-vector multiplication is much smaller than the errors caused by numerically solving $\bA \bx = \bb$ for $\bx$.
We can thus assess the accuracy of the three solvers by comparing the numerically computed $\bx$ to the ground truth $\bx_0$.

### Task &#x2F00;

First compare the outputs of matrix-vector multiplication $\bx \to \bA \bx$ using `double` and `long double` using the provided Rcpp functions.

**Questions:**
What is the relative difference in $\ell^2$-norm? 
How about the coordinate-wise relative differences?
Are the observed magnitudes of the differences what you'd expect?  

**Answers:** The relative difference in $\ell^2$-norm is 1.41e-30. The first 5 absolute value of coordinate-wise relative differences are 6.71e-15, 7.14e-15, 3.24e-15, -1.08e-15, -7.56e-16. The average of the coordinate-wise relative differences is 1.49e-16. The observed magnitudes of the differences is what I would expect. We expect these two functions would give extremely close solutions. Since the machine precision is $10^{-16}$, we expect the differences to be around this number.

```{r, eval=TRUE}
Rcpp::sourceCpp(file.path("src", "matvec_double.cpp"))
Rcpp::sourceCpp(file.path("src", "matvec_ldouble.cpp"))

set.seed(1918)

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
x <- rnorm(mat_size)

matvec_double_result <- matvec_double(A, x)
matvec_ldouble_result <- matvec_ldouble(A, x)
rel_diff <- (sqrt(sum((matvec_double_result-matvec_ldouble_result)^2)))/sqrt(sum(matvec_ldouble_result^2))
rel_diff
coord_diff <- (matvec_double_result-matvec_ldouble_result)/matvec_ldouble_result
head(coord_diff,10)
```

### Task &#x2F06;

Now randomly generate $\bA$ so that its condition number is $10^6$.
Then solve a positive-definite system $\bA \bx = \bb$ for $\bx$ using the three algorithms and compare their outputs to the ground truth $\bx_0$.

**Questions:**
Which algorithm appears to be more accurate than the others? 
Visually demonstrate your answer.   

**Answer:** From the graph, we can see that solve() is more accurate than the others.

```{r, eval=TRUE}
set.seed(1918)
cond_num <- 1e6

# Larger matrices could incur substantial computational time under base R BLAS
mat_size <- 1024L 

A <- rand_positive_def_matrix(mat_size, cond_num)
x <- rnorm(mat_size)

b <- matvec_ldouble(A, x)
method1 <- solve(A, b)
method_chol <- solve_via_cholesky(A, b)
method_qr <- qr.solve(A, b)

plot(method_qr-x, col="blue", ylab="error", pch=19, cex=0.5)
points(method_chol-x, col="red", pch=19, cex=0.5)
points(method1-x, col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))
```

### Task &#x4E09;

In Task &#x2F06;, we compared the three algorithms in one randomly generated example.
Now we consider a more systematic (though hardly comprehensive) comparison via repeated simulations.
We also vary the condition number of $\bA$ and assess whether the results would hold across varying degrees of ill-conditioning.

**Questions/To-do's:**

* Using the starter code provided, calculate various summary measures of the numerical errors.
* Integrate into the provided code one another (or more, if you like) meaningful metric(s) of your choice to summarize the numerical error.
* Visually explore how the three algorithms compare with each other in their accuracy. See if different error metrics tell different stories; they might or might not.
* Vary the condition number in the range $10^6 \sim 10^{12}$, e.g. by trying $10^6$, $10^9$, and $10^{12}$.
* Do you see any patterns in the numerical errors across the three algorithms, metrics, and/or condition numbers? Show some plots to support your conclusion.  

**Answers:** Within each condition number, across metrics, the points have the same patterns for each of the algorithm. For each metric and condition number, the plots show that QR decomposition generally has the largest errors, followed by cholesky decomposition and LU decomposition. As condition numbers decreases, relative errors for all three algorithms become smaller. 

**Note:** 
The QR solver will throw an error when the system is ill-condition enough that the numerical solution might not be very accurate. 
To force it to return the solution in any case, set `tol = .Machine$double.eps`.

```{r, eval=TRUE, fig.height=15, fig.width=8}
# Utility functions for bookkeeping simulation results.
source(file.path("R", "num_linalg_sim_study_helper.R"))

n_sim <- 32L
mat_size <- 512L
cond_num <- 1e6
metrics <- c("norm", "median", "five_percentile", "ninety_five_percentile", "average")
  # TODO: add another metric and modify the helper script accordingly

rel_error_list <- lapply(
  c("lu", "chol", "qr"), 
  function(method) pre_allocate_error_list(n_sim, metrics)
)

for (sim_index in 1:n_sim) {
  A <- rand_positive_def_matrix(mat_size, cond_num)
  x <- rnorm(mat_size) 
  b <- matvec_ldouble(A, x)
  x_approx <- list( 
    lu=solve(A, b),
    chol=solve_via_cholesky(A, b),
    qr=qr.solve(A, b, tol = .Machine$double.eps)
  )
  for (method in c("lu", "chol", "qr")) {
    rel_error <- lapply(
      metrics, 
      function (metric) calc_rel_error(x, x_approx[[method]], metric)
    )
    names(rel_error) <- metrics
    for (metric in names(rel_error)) {
      rel_error_list[[method]][[metric]][sim_index] <- rel_error[[metric]]
    }
  }
}

# TODO: visually compare errors
par(mfrow=c(5,1))
plot(rel_error_list[["qr"]][["norm"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["norm"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["norm"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["median"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["median"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["median"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["five_percentile"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["five_percentile"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["five_percentile"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["ninety_five_percentile"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["ninety_five_percentile"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["ninety_five_percentile"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["average"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["average"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["average"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))
```

```{r, eval=TRUE, fig.height=15, fig.width=8}
# Utility functions for bookkeeping simulation results.
source(file.path("R", "num_linalg_sim_study_helper.R"))

n_sim <- 32L
mat_size <- 512L
cond_num <- 1e9
metrics <- c("norm", "median", "five_percentile", "ninety_five_percentile", "average")
  # TODO: add another metric and modify the helper script accordingly

rel_error_list <- lapply(
  c("lu", "chol", "qr"), 
  function(method) pre_allocate_error_list(n_sim, metrics)
)

for (sim_index in 1:n_sim) {
  A <- rand_positive_def_matrix(mat_size, cond_num)
  x <- rnorm(mat_size) 
  b <- matvec_ldouble(A, x)
  x_approx <- list( 
    lu=solve(A, b),
    chol=solve_via_cholesky(A, b),
    qr=qr.solve(A, b, tol = .Machine$double.eps)
  )
  for (method in c("lu", "chol", "qr")) {
    rel_error <- lapply(
      metrics, 
      function (metric) calc_rel_error(x, x_approx[[method]], metric)
    )
    names(rel_error) <- metrics
    for (metric in names(rel_error)) {
      rel_error_list[[method]][[metric]][sim_index] <- rel_error[[metric]]
    }
  }
}

# TODO: visually compare errors
par(mfrow=c(5,1))
plot(rel_error_list[["qr"]][["norm"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["norm"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["norm"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["median"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["median"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["median"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["five_percentile"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["five_percentile"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["five_percentile"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["ninety_five_percentile"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["ninety_five_percentile"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["ninety_five_percentile"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["average"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["average"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["average"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))
```

```{r, eval=TRUE, fig.height=15, fig.width=8}
# Utility functions for bookkeeping simulation results.
source(file.path("R", "num_linalg_sim_study_helper.R"))

n_sim <- 32L
mat_size <- 512L
cond_num <- 1e12
metrics <- c("norm", "median", "five_percentile", "ninety_five_percentile", "average")
  # TODO: add another metric and modify the helper script accordingly

rel_error_list <- lapply(
  c("lu", "chol", "qr"), 
  function(method) pre_allocate_error_list(n_sim, metrics)
)

for (sim_index in 1:n_sim) {
  A <- rand_positive_def_matrix(mat_size, cond_num)
  x <- rnorm(mat_size) 
  b <- matvec_ldouble(A, x)
  x_approx <- list( 
    lu=solve(A, b),
    chol=solve_via_cholesky(A, b),
    qr=qr.solve(A, b, tol = .Machine$double.eps)
  )
  for (method in c("lu", "chol", "qr")) {
    rel_error <- lapply(
      metrics, 
      function (metric) calc_rel_error(x, x_approx[[method]], metric)
    )
    names(rel_error) <- metrics
    for (metric in names(rel_error)) {
      rel_error_list[[method]][[metric]][sim_index] <- rel_error[[metric]]
    }
  }
}

# TODO: visually compare errors
par(mfrow=c(5,1))
plot(rel_error_list[["qr"]][["norm"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["norm"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["norm"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["median"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["median"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["median"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["five_percentile"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["five_percentile"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["five_percentile"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["ninety_five_percentile"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["ninety_five_percentile"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["ninety_five_percentile"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))

plot(rel_error_list[["qr"]][["average"]], col="blue", ylab="error", pch=19, cex=0.5)
points(rel_error_list[["chol"]][["average"]], col="red", pch=19, cex=0.5)
points(rel_error_list[["lu"]][["average"]], col="black", pch=19, cex=0.5)
legend("right", legend = c("QR", "Chol", "LU"), col = c("blue", "red", "black"), pch = c(19,19,19))
```



