\documentclass{article}


\title{One reason to version control source files only}

\begin{document}

\maketitle

This is the first sentence.

This is sentence no.2.
This is sentence no.3.
This is sentence no.4.
This is sentence no.5.
This is sentence no.6.
This is sentence no.7.
This is sentence no.8.
This is sentence no.9.
This is sentence no.10.
This is sentence no.11.
This is sentence no.12.
This is sentence no.13.
This is sentence no.14.
This is sentence no.15.
This is sentence no.16.
This is sentence no.17.
This is sentence no.18.
This is sentence no.19.
This is sentence no.20.
This is sentence no.21.
This is sentence no.22.
This is sentence no.23.
This is sentence no.24.
This is sentence no.25.
This is sentence no.26.
This is sentence no.27.
This is sentence no.28.
This is sentence no.29.
This is sentence no.30.
This is sentence no.31.
This is sentence no.32.
This is sentence no.33.
This is sentence no.34.
This is sentence no.35.
This is sentence no.36.
This is sentence no.37.
This is sentence no.38.
This is sentence no.39.
This is sentence no.40.
This is sentence no.41.
This is sentence no.42.
This is sentence no.43.
This is sentence no.44.
This is sentence no.45.
This is sentence no.46.
This is sentence no.47.
This is sentence no.48.
This is sentence no.49.
This is sentence no.50.
This is sentence no.51.
This is sentence no.52.
This is sentence no.53.
This is sentence no.54.
This is sentence no.55.
This is sentence no.56.
This is sentence no.57.
This is sentence no.58.
This is sentence no.59.
This is sentence no.60.
This is sentence no.61.
This is sentence no.62.
This is sentence no.63.
This is sentence no.64.
This is sentence no.65.
This is sentence no.66.
This is sentence no.67.
This is sentence no.68.
This is sentence no.69.
This is sentence no.70.
This is sentence no.71.
This is sentence no.72.
This is sentence no.73.
This is sentence no.74.
This is sentence no.75.
This is sentence no.76.
This is sentence no.77.
This is sentence no.78.
This is sentence no.79.
This is sentence no.80.
This is sentence no.81.
This is sentence no.82.
This is sentence no.83.
This is sentence no.84.
This is sentence no.85.
This is sentence no.86.
This is sentence no.87.
This is sentence no.88.
This is sentence no.89.
This is sentence no.90.
This is sentence no.91.
This is sentence no.92.
This is sentence no.93.
This is sentence no.94.
This is sentence no.95.
This is sentence no.96.
This is sentence no.97.
This is sentence no.98.
This is sentence no.99.
This is sentence no.100.
\end{document}